---
title: "Github_SCC"
output: html_document
date: "2025-02-27"
---

Import packages
```{r}
library(ape, quietly = TRUE)
library(caret, quietly = TRUE)
library(cowplot, quietly = TRUE)
library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(ggpubr, quietly = TRUE)
library(grid, quietly = TRUE)
library(ggnewscale, quietly = TRUE)
library(ggtree, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(knitr, quietly = TRUE)
library(magrittr, quietly = TRUE)
library(patchwork, quietly = TRUE)
library(phyloseq, quietly = TRUE)
library(pROC, quietly = TRUE)
library(randomcoloR, quietly = TRUE)
library(randomForest, quietly = TRUE)
library(RColorBrewer, quietly = TRUE)
library(readr)
library(readxl, quietly = TRUE)
library(reshape2, quietly = TRUE)
library(rhdf5)
library(Rtsne)
library(scales, quietly = TRUE)
library(SIAMCAT, quietly = TRUE)
library(tidyr, quietly = TRUE)
library(vegan, quietly = TRUE)
rm(list=ls())
```


Base theme for plotting

```{r}
base_theme <- theme(
    text = element_text(size = 14),
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 15),
    plot.title = element_text(size = 17),
    legend.text = element_text(size = 16),   # Legend label text size
    legend.title = element_text(size = 17),
    strip.text = element_text(size = 17),
    plot.margin = unit(c(0,0,0,0), "cm")
  )
```


This study analyzed a cohort of 30 subjects, comprising 15 SOTRs and 15 non-transplant subjects subdivided into Low-Risk (n=9) and High-Risk (n=6) subgroups based on cutaneous SCC history (Supplementary Table 1). The Low-Risk group included healthy subjects not on immunosuppressive medications and with two or fewer SCCs (including squamous cell carcinoma in situ [SCCIS]). In contrast, the High-Risk group consisted of non-immunocompromised subjects with more than two SCCs. The mean SCC count was highest in High-Risk subjects (11.8), followed by SOTRs (8.7), and Low-Risk subjects (0.33) (Supplementary Figure 1). The High-Risk group tended to be older, with a median age of 78.5 years compared to 63 years in the Low-Risk group and 68 years in the SOTRs. A positive correlation between age and SCC counts was observed (Supplementary Figure 1).

```{r}
metadata <- read.csv("SuppTable1.csv", stringsAsFactors = FALSE)

nrow(metadata) #number of samples
unique(metadata$site) #Ct is control

summary_stats <- metadata %>%
  filter(visit %in% c("one_visit", "tp_2")) %>%            #don't want to include multiple visits per subject
  distinct(subject_id, Patient_group, Age, SCC_count) %>%     
  group_by(Patient_group) %>%
  summarise(
    n_subjects    = n_distinct(subject_id),
    median_age    = median(Age, na.rm = TRUE),
    mean_SCC_count = mean(SCC_count, na.rm = TRUE)
  )

# View the summary
print(summary_stats)
```


Figure 1
```{r}
metadata_indivsub <- metadata %>%
  filter(visit %in% c("one_visit", "tp_2")) %>%  # only include one visit per person
  distinct(subject_id, Age, Patient_group, SCC_count)
metadata_indivsub$Patient_group=factor(metadata_indivsub$Patient_group, levels=c("Low-Risk SCC", "High-Risk SCC", "SOTR"))


# Create a dot plot for Age vs. SCC_count colored by Patient_group
plot_age_scc <- ggplot(metadata_indivsub, aes(x = Age, y = SCC_count, color = Patient_group)) +
  geom_point(size = 2) +
  labs(
    x = "Age",
    y = "Number of SCCs",
    color = "Patient group"
  ) + theme_bw() +
  base_theme +
  theme(legend.position = "right") +
  scale_color_manual(values = c("Low-Risk SCC" = "#33a02c",
                                "High-Risk SCC" = "#1f78b4",
                                "SOTR" = "#e31a1c"))

plot_age_scc

# Save the plot to a file
ggsave(filename = "final_figures/fig1.png", plot = plot_age_scc, width=8, height=4)
```


Figure 2
```{r}
# Define variables
var.name <- "20230806_standard++_0.5_bracken"

# Read in the phyloseq object, normalize counts, melt to long format,
# filter out unwanted sites, and clean up column values
physeq_df_orig <- readRDS(paste0(var.name, ".rds")) %>%
 transform_sample_counts(function(x) x/sum(x)) %>%
 psmelt() %>%
 filter(!site %in% c("Tw", "Um", "Fw")) %>% #Don't have data for all subjects across these sites
 mutate(Family = gsub("f__", "", Family))

# Identify the top 6 phyla by total Abundance
target_phyla <- physeq_df_orig %>%
  group_by(Phylum) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  slice_max(Abundance, n = 6) %>%
  pull(Phylum)

# For each Phylum, select the top 3 families (by summed Abundance)
top_families <- physeq_df_orig %>%
  group_by(Phylum, Family) %>%
  summarise(total_sum = sum(Abundance), .groups = "drop") %>%
  group_by(Phylum) %>%
  slice_max(total_sum, n = 3) %>%
  filter(Phylum %in% target_phyla)

# Define a color palette and an ordered list of families
palette <- c("white",
  "#31A354", "#74C476", "#BAE4B3", "white",           # Green shades
  "#045a8d", "#2b8cbe", "#74a9cf", "white",             # Blue shades
  "#a50f15", "#de2d26", "#fb6a4a", "white",             # Red shades
  "#756bb1", "#9e9ac8", "#bcbddf", "white",             # Purple shades
  "#fbb4b9", "#f768a1", "#fc4c70", "white",             # Pink shades
  "#fe9929", "#fec44f", #"white",                        # Orange shades
  "gray"
)
names(palette) <- c("Actinomycetota", "Propionibacteriaceae", "Corynebacteriaceae",
                    "Micrococcaceae", "Bacillota", "Staphylococcaceae", "Streptococcaceae",
                    "Peptoniphilaceae", "Bacteroidota", "Prevotellaceae", "Bacteroidaceae",
                    "Weeksellaceae", "Pseudomonadota", "Moraxellaceae", "Xanthomonadaceae",
                    "Enterobacteriaceae", "Basidiomycota", "Malasseziaceae", "Pucciniaceae",
                    "Ustilaginaceae", "Cossaviricota", "Papillomaviridae", "Polyomaviridae",
                    "Other")

families_list <- c("Propionibacteriaceae", "Corynebacteriaceae", "Micrococcaceae",
                   "Staphylococcaceae", "Streptococcaceae", "Peptoniphilaceae",
                   "Prevotellaceae", "Bacteroidaceae", "Weeksellaceae", "Moraxellaceae",
                   "Xanthomonadaceae", "Enterobacteriaceae", "Malasseziaceae",
                   "Pucciniaceae", "Ustilaginaceae", "Papillomaviridae",
                   "Polyomaviridae")

# Prepare plot data by summarizing Abundance for each combination of metadata,
# and recoding Family as a factor with specified levels (top families vs. "Other")
physeq_df_orig <- physeq_df_orig %>%
  group_by(site, subject_id, Phylum, Family, Patient_group, cancer_wart, Wart, SCC) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  mutate(Family = factor(ifelse(Family %in% top_families$Family, Family, "Other"), levels = c(families_list, "Other")))


# For each unique site and each subject (with associated metadata),
# add a row where Phylum and Family are "Other" and Abundance is 0.
subject_info <- physeq_df_orig %>%
  distinct(subject_id, Patient_group, cancer_wart, Wart, SCC)
site_list <- tibble(site = unique(physeq_df_orig$site))
other_rows <- site_list %>%
  crossing(subject_info) %>%
  mutate(Phylum = "Other",
    Family = "Other",
    Abundance = 0)
# Combine the original data with the additional rows
physeq_df_final <- bind_rows(physeq_df_orig, other_rows)

physeq_df_final$Family <- factor(physeq_df_final$Family, levels = c(families_list, "Other")) #????


physeq_df_final$subject_id = factor(physeq_df_final$subject_id , levels = c("H1", "H2", "H3", "H4", "H5", "H6", "L1", "L2", "L3", "L4", "L5", "L6", "L7", "L8", "L9", "T1", "T2","T3", "T4","T5", "T6", "T7", "T8", "T9", "T10", "T11", "T12", "T13", "T14", "T15"))
# Subset data by site using filter instead of subset
physeq_fh <- filter(physeq_df_final, site == "Fh")
physeq_ac <- filter(physeq_df_final, site == "Ac")
physeq_pc <- filter(physeq_df_final, site == "Pc")

# Function to create abundance and clinical history plots for a given site
plot_abundance <- function(data, site_code, palette, tile) {
  # Map site codes to descriptive titles
  plot_title <- case_when(
    site_code == "Pc" ~ "Popliteal crease",
    site_code == "Fh" ~ "Forehead", 
    site_code == "Ac" ~ "Antecubital crease",
    TRUE ~ site_code
  )
  
  
  # Main abundance bar plot
  p1 <- ggplot(data, aes(x = subject_id, y = Abundance, fill = Family)) +
    geom_bar(stat = "identity", position = "fill") +
    scale_fill_manual(values = palette) +
    scale_y_continuous(expand = c(0, 0)) +
    facet_grid(~Patient_group, scales = "free", space = "free") +
    labs(title = plot_title, x = NULL, y = NULL) + theme_bw()+
    base_theme +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      legend.position = "none"
    )
  
  # Clinical history tile plot (for Wart and SCC)
  tile_data <- data %>%
    distinct(subject_id, Patient_group, Wart, SCC) %>%
    pivot_longer(cols = c(Wart, SCC), names_to = "condition") %>%
    mutate(color = ifelse(value == "Yes", tolower(condition), "white"))
 # tile_data$subject_id = factor(physeq_df_final$subject_id , levels = c("H1", "H2", "H3", "H4", "H5", "H6", "L1", "L2", "L3", "L4", "L5", "L6", "L7", "L8", "L9", "T1", "T2","T3", "T4","T5", "T6", "T7", "T8", "T9", "T10", "T11", "T12", "T13", "T14", "T15"))

  
  p2 <- ggplot(tile_data, aes(x = subject_id, y = condition, fill = color)) +
    geom_tile(color = "black", size = 0.1) +
    scale_fill_manual(
      values = c(wart = "blue", scc = "green", white = "white"),
      labels = c(wart = "Wart: Yes", scc = "SCC: Yes", white = "No"),
      name = "Clinical history:"
    ) +
    facet_grid(~Patient_group, scales = "free", space = "free") +
    labs(x = NULL, y = NULL) +
    theme_void() +
    base_theme +
    theme(
      strip.text = element_blank(),
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
      legend.position = "right"
    )
  
  # Combine plots and extract legends based on the tile argument
  if (tile) {
    main_plot <- p1 / (p2 + theme(legend.position = "none")) +
      plot_layout(heights = c(10, 3))
    
    legends <- plot_grid(
      get_legend(p1 + theme(legend.position = "right")),
      get_legend(p2 + theme(legend.position = "right")),
      ncol = 1,
      rel_heights = c(9, 3)
    )
  } else {
    main_plot <- p1
    legends <- plot_grid(
      get_legend(p1 + theme(legend.position = "right")),
      ncol = 1,
      rel_heights = c(10, 3)
    )
  }
  
  list(main_plot = main_plot, legend_plot = legends)
}

# Generate plots for each site using the refined function
plot_fh <- plot_abundance(physeq_fh, "Fh", palette, tile = FALSE)
plot_ac <- plot_abundance(physeq_ac, "Ac", palette, tile = TRUE)
plot_pc <- plot_abundance(physeq_pc, "Pc", palette, tile = FALSE)

# Combine the individual plots into one final composite plot
final_plota <- plot_grid(
  plot_grid(
    NULL,
    plot_grid(
      plot_pc$main_plot,
      NULL,
      plot_fh$main_plot,
      NULL,
      plot_ac$main_plot,
      NULL,
      nrow = 6, rel_heights = c(0.3, 0.03, 0.3, 0.03, 0.3, 0.03)
    ),
    NULL,
    plot_ac$legend_plot,
    nrow = 1,
    rel_widths = c(0.03, 0.79, 0.02, 0.17)
  ) +
    draw_label("Subjects", x = 0.4, y = 0.01, size = 18) +
    draw_label("Relative Abundance", x = 0.01, y = 0.5, angle = 90, size = 18)) 

final_plota
```

```{r}
# Load data and preprocess
ps <- readRDS(paste0(var.name, ".rds"))

taxa_abundance_table <- ps %>%
  tax_glom(taxrank = "Kingdom") %>%
  transform_sample_counts(function(x) x/sum(x)) %>%
  psmelt() %>%
  filter(!Kingdom %in% "k__Archaea") %>%
  mutate(Kingdom = gsub("k__", "", gsub("Eukaryota", "Eukaryotes", Kingdom)),  
         `Immune status` = factor(Patient_group, levels = c("Low-Risk SCC","High-Risk SCC","SOTR")),
         Kingdom = factor(Kingdom, levels = unique(Kingdom)))


taxa_abundance_table_limit=taxa_abundance_table%>%filter(site %in% c("Ac", "Pc", "Fh")) #subset sites

finalb <- ggplot(taxa_abundance_table_limit, aes(x = site, y = Abundance)) +
  geom_boxplot(aes(fill = `Immune status`), alpha = 0.8, outlier.shape = NA) +
  geom_point(aes(color = `Immune status`), alpha = 0.7, 
             position = position_jitterdodge(0.2)) +
  scale_fill_manual(values = c("Low-Risk SCC" = "#33a02c", "High-Risk SCC" = "#1f78b4", "SOTR" = "#e31a1c")) +
  scale_color_manual(values = c("Low-Risk SCC" = "#33a02c", "High-Risk SCC" = "#1f78b4", "SOTR" = "#e31a1c")) +
  labs(y = "Relative Abundance", 
       x = "Site", 
       fill = "Patient group", 
       color = "Patient group") + theme_bw()+
  base_theme +
  theme(legend.position = "bottom") +
  guides(color = "none", fill = guide_legend(nrow = 1)) +
  facet_grid(. ~ Kingdom)

finalb
```

```{r}
fig2 <- plot_grid(final_plota, NULL, finalb, nrow=3, rel_heights = c(0.73,0.02, 0.25), labels = c("A","","B"), label_size = 20)
ggsave("final_figures/fig2.png", fig2, width=13, height=15)
```


Supp fig 1 - Full set of sites from fig 2b
```{r}
# Compare means by immune status within each Kingdom across all sites
compare_means(Abundance ~ Patient_group, data = taxa_abundance_table, 
              group.by = "Kingdom", ref.group = "Low-Risk SCC",
              method = "t.test")

comparisons=list(c("Low-Risk SCC", "SOTR"),c("Low-Risk SCC", "High-Risk SCC"),c("High-Risk SCC", "SOTR"))

suppfig1 <- taxa_abundance_table %>%
  ggplot(aes(`Immune status`, Abundance, fill = `Immune status`)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(color = `Immune status`), size = 1.75, 
            position = position_dodge(0.75)) +
  facet_grid(Kingdom ~ site, scales = "free_x", space = "free_x") +
  stat_compare_means(
    comparisons = comparisons, 
    method = "t.test",
    size = 5,
    label = "p.signif"
  ) +
  scale_fill_manual(values = c("Low-Risk SCC" = "#33a02c", "High-Risk SCC" = "#1f78b4", "SOTR" = "#e31a1c")) +
  scale_color_manual(values = c("Low-Risk SCC" = "#33a02c", "High-Risk SCC" = "#1f78b4", "SOTR" = "#e31a1c")) +
  labs(y = "Relative Abundance", x = "Patient group", fill = "Patient group") +
  theme_bw() +
  base_theme +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 18),
    legend.position="bottom"
  ) + guides(color = "none") + scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))

suppfig1

ggsave("final_figures/suppfig1.png", suppfig1, width=12, height=12)
```




```{r}
# Define parameters and initialize lists
titles <- c("SOTR vs Low Risk", "High Risk vs Low Risk")
sites <- c("Ac", "Pc")
immune_status_list <- list(c("SOTR", "Low-Risk SCC"), c("High-Risk SCC", "Low-Risk SCC"))
comparison_labels <- c("Transplant - Low risk", "High risk - Low risk")
physeq_list <- siamcat_list <- associations_list <- list()

for (i in 1:2) {
  
  # 1. Read and subset phyloseq object
  physeq <- readRDS(paste0(var.name, ".rds"))
  physeq <- subset_samples(physeq, (site %in% sites) & (Patient_group %in% immune_status_list[[i]]))
  
  # 2. Process taxonomy and OTU tables
  tax_mat <- as.data.frame(tax_table(physeq)) %>%
    mutate(Genus = gsub("g__", "", Genus),
           Species = gsub("s__", "", Species))
  rownames(tax_mat) <- paste(tax_mat$Genus, tax_mat$Species, sep = " ")
  otutable <- physeq@otu_table
  colnames(otutable) <- rownames(tax_mat)
  
  # 3. Read and subset metadata
  metadata <- read.csv("SuppTable1.csv", stringsAsFactors = FALSE)
  selected_columns <- c("Sample", "Patient_group")
  metadata <- metadata[, colnames(metadata) %in% selected_columns]
  rownames(metadata) <- metadata$Sample
  
  # 4. Recreate phyloseq object and normalize
  physeq <- phyloseq(otutable, tax_table(as.matrix(tax_mat)), sample_data(metadata[rownames(otutable),]))
  physeq_rel <- transform_sample_counts(physeq, function(x) x / sum(x))
  
  # 5. Prepare data for SIAMCAT
  feature_matrix <- t(otu_table(physeq_rel))
  meta_data <- as.data.frame(sample_data(physeq_rel))
  labels <- create.label(meta = meta_data, label = "Patient_group",
                         case = immune_status_list[[i]][1],
                         control = immune_status_list[[i]][2])
  siamcat_obj <- siamcat(feat = feature_matrix, label = labels, meta = meta_data)
  siamcat_obj <- validate.data(siamcat_obj)
  
  # 6. Feature selection and association testing
  siamcat_obj <- filter.features(siamcat_obj, filter.method = 'abundance', cutoff = 0.05)
  siamcat_obj <- check.associations(siamcat_obj,
                                    log.n0 = 1e-05,
                                    test = 'wilcoxon',
                                    alpha = 0.05,
                                    mult.corr = "fdr",
                                    feature.type = 'filtered')
  
  # 7. Process associations
  associations <- siamcat_obj@associations$assoc.results
  associations <- associations[associations$p.adj < 0.05, ]
  associations$comparison <- comparison_labels[i]
  associations$species <- rownames(associations)
  
  # Store objects for later use
  physeq_list[[i]] <- physeq
  siamcat_list[[i]] <- siamcat_obj
  associations_list[[i]] <- associations
  
  # Combine associations from all iterations for plotting
  associations_all <- do.call(rbind, associations_list)
  associations_all$genus <- gsub("_s__.*", "", associations_all$species)
  
  # 8. Further process associations for additional plots
  associations_all$Species <- rownames(associations_all)
  associations_all$p.adj <- as.numeric(as.vector(associations_all$p.adj))
  levels_order <- associations_all$Species[order(associations_all$fc, decreasing = TRUE)]
  associations_all$Species <- factor(associations_all$Species, levels = levels_order)
  associations_all$color <- if_else(associations_all$fc < 0, 'cornflowerblue', '#ffc125')
  associations_all$log10padj <- -log10(associations_all$p.adj)
  
  # 9. Process relative abundance data
  relabund <- psmelt(physeq_rel)
  relabund <- relabund[relabund$OTU %in% levels_order, ]
  relabund$Species <- factor(relabund$OTU, levels = levels_order)
  relabund$Abundance <- log10(relabund$Abundance)
  relabund$`Immune status` <- relabund$Patient_group
  
  # 10. Define common theme for additional plots
  common_theme <- theme_bw() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.grid.major.y = element_blank(),
          panel.grid.minor.y = element_blank(),
          panel.grid.major.x = element_line(color = "gray", size = 0.5),
          panel.background = element_blank(),
          axis.line = element_line(color = "black"))
  
  # Create individual plots for pval, auc, and fc
  pval <- ggplot(associations_all, aes(x = Species, y = log10padj)) +
    geom_bar(stat = "identity", fill = "grey") +
    coord_flip() +
    common_theme +
    xlab("") + ylab("-log10(p.adj)") + base_theme
  
  auc <- ggplot(associations_all, aes(x = Species, y = auc, ymin = auc.ci.l, ymax = auc.ci.h)) +
    geom_linerange(color = "grey", size = 1, alpha = 0.5) +
    geom_point(color = "grey", size = 3) +
    coord_flip() +
    ylim(0, 1) +
    common_theme +
    xlab("") + ylab("AUC") + base_theme
  
  fc <- ggplot(associations_all, aes(x = Species, y = fc, fill = color)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_manual(values = c('grey', 'grey')) +
    common_theme +
    theme(legend.position = "none") +
    xlab("") + ylab("Fold Change") + base_theme
  
  # Relative abundance boxplot
  relabd <- ggplot(relabund, aes(x = Species, y = Abundance, fill = `Immune status`, color = `Immune status`)) +
    geom_boxplot() +
    geom_jitter(alpha = 0.2, size = 1) +
    coord_flip() +
    labs(x = "Species", y = "Relative Abundance (log10)", fill = "Patient group", color = "Patient group") +
    theme_bw() + base_theme +
    theme(panel.grid.major.x = element_line(color = "gray", size = 0.5),
          panel.background = element_blank(),
          axis.line = element_line(color = "black"),
          legend.position = "bottom") +
    scale_color_manual(values = c("#33a02c", "#1f78b4", "#e31a1c"),
                       breaks = c("Low-Risk SCC", "High-Risk SCC", "SOTR")) +
    scale_fill_manual(values = c("#33a02c", "#1f78b4", "#e31a1c"),
                      breaks = c("Low-Risk SCC", "High-Risk SCC", "SOTR")) +
    guides(fill = guide_legend(nrow = 2, byrow = TRUE))
  
  # Extract legend and combine plots
  legend <- get_legend(relabd + theme(legend.position = "top"))
  legend <- plot_grid(legend, NA, ncol = 2)
  
  variable_name <- paste("fig3a", i, sep = "_")
  title_plot <- ggdraw() + draw_label(titles[i], size = 18)
  
  combined_plot <- plot_grid(title_plot,
    plot_grid(relabd + theme(legend.position = "none"), NULL, pval,NULL, fc,NULL, auc + theme(legend.position = "none"), NULL,
              ncol = 8, rel_widths = c(0.39, 0.01, 0.19,0.01, 0.19,0.01, 0.19, 0.01)),
    ncol = 1, rel_heights = c(0.1, 0.9)
  )
  
  assign(variable_name, combined_plot)
}

# Combine the two figures vertically
fig3a <- plot_grid(fig3a_1, fig3a_2, nrow = 2, rel_heights = c(0.4, 0.6))
print(fig3a)
```


Relative abundance of HPV and wart status
```{r}
# Read phyloseq object
ps <- readRDS(file.path(paste0(var.name, ".rds")))

# Aggregate data at the Family level and compute relative abundances
physeq_rel <- ps %>%
  tax_glom(taxrank = "Family")
  
physeq_rel <- transform_sample_counts(physeq_rel, function(x) x / sum(x))


# Melt data and filter for Papillomaviridae in selected sites/visits;
# clean up the "History.of.Warts.as.Adults" variable
taxa_abundance_table <- psmelt(physeq_rel) %>%
  filter(Family == "f__Papillomaviridae",
         site %in% c("Ac", "Pc"),
         visit %in% c("tp_1", "one_visit")) %>%
  mutate(
    History.of.Warts.as.Adults = gsub(" .*", "", History.of.Warts.as.Adults),
    History.of.Warts.as.Adults = recode(History.of.Warts.as.Adults,
                                         "YES" = "Warts present",
                                         "NO"  = "Warts not\npresent")
  )

# Compute average abundance by grouping variables
average_abundance <- taxa_abundance_table %>%
  group_by(site, subject_id, History.of.Warts.as.Adults, Patient_group) %>%
  summarize(Abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")

# Define comparisons for t-test
comparisons <- list(c("Warts not\npresent", "Warts present"))

# Create the plot
fig3b <- ggplot(average_abundance, aes(x = History.of.Warts.as.Adults, y = Abundance)) +
  theme_bw() +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  base_theme +
  geom_jitter(aes(color = Patient_group, size = 2), width = 0.2, size = 2) +
  stat_compare_means(method = "t.test", label = "p.signif", comparisons = comparisons, size=10) +
  xlab("History of warts as adults") +
  ylab("Relative abundance of\nPapillomaviridae") +
  labs(color = "Patient group") +
  scale_color_manual(values = c("#33a02c", "#1f78b4", "#e31a1c"),
                     breaks = c("Low-Risk SCC", "High-Risk SCC", "SOTR")) + theme(legend.position="bottom") + scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) + guides(size = guide_legend(override.aes = list(size = 3)), color = guide_legend(nrow = 2))


# Extract legend (if needed)
legend2 <- plot_grid(get_legend(fig3b + guides(color = guide_legend(nrow = 1)) + theme(legend.position = "bottom")))

fig3b
```




HPV tree

```{r}
heatmapData = readRDS("HPV_tree/met_hpv_git.RDS")
heatmapData$`Genome source`=factor(heatmapData$`Genome source`, levels=c("Low-Risk SCC", "High-Risk SCC", "SOTR", "NCBI"))
table(heatmapData$`Genome source`)

tree_trim <- read.tree("HPV_tree/trim.tree")

alpha=heatmapData$taxa[which(heatmapData$ScientificName=="Alphapapillomavirus")]
alpha_val=getMRCA(tree_trim, tip=alpha) #191
beta=heatmapData$taxa[which(heatmapData$ScientificName=="Betapapillomavirus")]
beta_val=getMRCA(tree_trim, tip=beta) #206
gamma=heatmapData$taxa[which(heatmapData$ScientificName=="Gammapapillomavirus")]
gamma_val=getMRCA(tree_trim, tip=gamma) #270
delta=heatmapData$taxa[which(heatmapData$ScientificName=="Deltapapillomavirus")]
delta_val=getMRCA(tree_trim, tip=delta) #182
epsilon=heatmapData$taxa[which(heatmapData$ScientificName=="Epsilonpapillomavirus")]
epsilon_val=getMRCA(tree_trim, tip=epsilon) #188


p2 <- ggtree(tree_trim, layout='circular') %<+% heatmapData +
  geom_tippoint(aes(color = `Genome source`)) +  # Adjust size as needed for visibility
  scale_color_manual(values = c("Low-Risk SCC" = "#33a02c", "High-Risk SCC" = "#1f78b4", "SOTR" = "#e31a1c", "NCBI"="grey"))+
  geom_cladelab(node=alpha_val, label="Alphapapillomavirus", fill="#CFABDE", geom = "label", barcolor="#CFABDE",align=TRUE, barsize=2,offset.text=0.4, offset =0.2, fontsize = 7) + 
    geom_cladelab(node=beta_val, label="Betapapillomavirus", fill= "#B3D9D9", geom = "label", barcolor= "#B3D9D9",align=TRUE, barsize=2,offset.text=0.4, offset =0.2, fontsize = 7) +  
    geom_cladelab(node=gamma_val, label="Gammapapillomavirus", fill= "#F5CEAB", geom = "label", barcolor= "#F5CEAB",align=TRUE, barsize=2,offset.text=0.4, offset =0.2, fontsize = 7) +  
      geom_cladelab(node=delta_val, label="Deltapapillomavirus", fill="#FFD1DC", geom = "label", barcolor="#FFD1DC",align=TRUE, barsize=2,offset.text=0.4, offset =0.2, fontsize = 7) +
      geom_cladelab(node=epsilon_val, label="Epsilonpapillomavirus", fill="#B99C85", geom = "label", barcolor="#B99C85", align=TRUE,barsize=2,offset.text=0.4, offset =0.2, fontsize = 7)+
  theme(
      legend.text = element_text(size = 16),
      legend.title = element_text(size = 18),
      legend.position = "bottom",
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      legend.background = element_rect(fill = "white", color = NA)
          ) + guides(size = guide_legend(override.aes = list(size = 3)))

p2
```
  


```{r}
# fig2_2<-plot_grid(fig2a,NULL, fig2c,ncol=3, labels=c("","B"),rel_widths = c(0.75,0.02, 0.23), label_size = 20)
# fig2_2<-plot_grid(fig2_2,legend2, nrow=2, labels=c("",""), rel_heights = c(0.9,0.1), label_size = 20)
# fig2_2<-plot_grid(fig2_2,p2,nrow=2,labels=c("","C"), rel_heights = c(0.4,0.6), label_size = 20)
# fig2_2


fig3a2<-plot_grid(fig3a, legend2, nrow=2, labels=c("A",""), rel_heights = c(0.9,0.1), label_size = 20)
fig3_2<-plot_grid(fig3b, NULL, p2, NULL, ncol=4, labels=c("B","", "C",""), rel_widths = c(0.28, 0.05,0.7,0.05), label_size = 20)
fig3<-plot_grid(fig3a2, NULL, fig3_2,nrow=3, rel_heights=c(0.45,0.03, 0.52), label_size = 20)
fig3

ggsave("final_figures/fig3.png", fig3, width=17, height=16)

```



Suppfig2
```{r}
#############################
### Suppfig2a: Eukaryotes ###
#############################

# Define taxa level and number of colors for eukaryotic genera
taxa_level <- "Genus"
n.colors <- 3

# Load the phyloseq object and process the data
physeq_orig <- readRDS(paste0(var.name, ".rds"))
physeq_rel <- transform_sample_counts(physeq_orig, function(x) x / sum(x))
physeq_df <- psmelt(physeq_rel)

# Set every genus where Kingdom is not "k__Eukaryota" to "Other", subset to site "Fh",
# and remove the "g__" prefix from Genus names
physeq_df$Genus[ which(!(physeq_df$Kingdom == "k__Eukaryota")) ] <- "Other"
physeq_df <- physeq_df[physeq_df$site == "Fh", ]
physeq_df$Genus <- gsub("g__", "", physeq_df$Genus)


# Identify the top genera based on total abundance (n.colors + 1) and reassign the rest to "Other"
top_genera <- physeq_df %>%
  group_by(Genus) %>%
  summarise(Abundance = sum(Abundance)) %>%
  arrange(desc(Abundance)) %>%
  slice_head(n = n.colors + 1) %>%
  pull(Genus)

physeq_df <- physeq_df %>%
  mutate(Genus = ifelse(Genus %in% top_genera, Genus, "Other"))

# Aggregate data by site, subject_id, Genus, and Patient_group
physeq_df <- physeq_df %>%
  group_by(site, subject_id, Genus, Patient_group) %>%
  summarise(Abundance = sum(Abundance), .groups = 'drop')

# Create a color palette: use a Brewer palette for top genera and add grey for "Other"
colors <- brewer.pal(n.colors, "Set2")
colors <- c(colors, "grey")
genera_colors <- setNames(colors, c(top_genera[ which(top_genera != "Other") ], "Other"))

physeq_df$subject_id = factor(physeq_df$subject_id , levels = c("H1", "H2", "H3", "H4", "H5", "H6", "L1", "L2", "L3", "L4", "L5", "L6", "L7", "L8", "L9", "T1", "T2","T3", "T4","T5", "T6", "T7", "T8", "T9", "T10", "T11", "T12", "T13", "T14", "T15"))

# Create the faceted bar plot (one panel per Patient_group)
suppfig2a <- ggplot(physeq_df, aes(x = subject_id, y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = genera_colors, guide = guide_legend(nrow = 2)) +
  labs(
    x = "Subjects",
    y = "Relative Abundance",
    fill = taxa_level
  ) +
  facet_wrap(~ Patient_group, ncol = 1, scales="free") +  theme_bw() + 
  base_theme

##############################
### Suppfig2b: Viruses      ###
##############################

# Load the phyloseq object and process the data for viruses
physeq_orig <- readRDS(paste0(var.name, ".rds"))
physeq_rel <- transform_sample_counts(physeq_orig, function(x) x / sum(x))
physeq_df <- physeq_rel %>%
  psmelt() %>%
  mutate(Genus = gsub("^g__", "", Genus)) %>%
  filter(
    site %in% c("Ac", "Pc"),
    Genus %in% c("Alphapolyomavirus", "Betapapillomavirus", "Gammapapillomavirus")
  ) %>%
  group_by(Sample, site, subject_id, visit, Genus, Patient_group, SCC_count) %>%
  summarise(Abundance = mean(Abundance), .groups = 'drop')

# Create the faceted scatter plot by virus Genus with significance comparisons
suppfig2b <- ggplot(physeq_df, aes(x = Patient_group, y = Abundance, color = Patient_group)) +
  geom_point(size = 1.75, position = position_dodge(width = 0.75)) +
  facet_wrap(~ Genus, nrow = 3) +
  stat_compare_means(
    comparisons = list(c("Low-Risk SCC", "SOTR"), c("Low-Risk SCC", "High-Risk SCC")),
    method = "t.test",
    size = 5,
    label = "p.signif"
  ) +
  scale_color_manual(values = c("Low-Risk SCC" = "#33a02c", "High-Risk SCC" = "#1f78b4", "SOTR" = "#e31a1c"),guide = guide_legend(nrow = 2)) +
  labs(
    x = "Patient Group",
    y = "Relative Abundance",
    color = "Patient Group") + theme_bw() + 
  base_theme +scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))



plot_grid(
  suppfig2a+theme(legend.position="bottom"), NULL, suppfig2b+theme(legend.position="bottom"),
  labels = c("A", "B"),
  ncol = 3,               
  align = "hv",           
  rel_widths = c(0.57, 0.02, 0.41)
)
ggsave("final_figures/suppfig2.png",width = 13, height=10)
```